---
title: "Lesson 17: Machine Learning with tidymodels"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
```

## Learning Objectives

By the end of this lesson, you will be able to:

1. Split data into training and testing sets
2. Preprocess data with recipes
3. Specify and train machine learning models
4. Use cross-validation for model evaluation
5. Tune hyperparameters systematically
6. Compare model performance
7. Make predictions on new data
8. Identify important features
9. Apply ML to biological classification and regression problems

## Why tidymodels for Biology?

Machine learning is increasingly important for:
- **Classification**: Disease status, cell type identification
- **Prediction**: Drug response, protein function
- **Feature selection**: Which genes/proteins matter most?
- **Pattern discovery**: Clustering patient subtypes
- **High-dimensional data**: Genomics, proteomics, imaging

tidymodels provides a consistent, tidyverse-compatible framework!

## Setup

```{r load-packages}
library(tidyverse)
library(tidymodels)  # Meta-package: includes parsnip, recipes, rsample, etc.
library(ranger)      # Random forest engine
library(glmnet)      # Regularized regression
library(vip)         # Variable importance plots

set.seed(123)
```

## The tidymodels Framework

```{r tidymodels-overview, eval=FALSE}
# tidymodels includes:
# - rsample:   Data splitting and resampling
# - recipes:   Feature engineering and preprocessing
# - parsnip:   Model specification
# - workflows: Combine recipes and models
# - tune:      Hyperparameter tuning
# - yardstick: Model evaluation metrics
# - broom:     Tidy model outputs
```

## Example Dataset: Predicting Disease Status

```{r create-data}
# Simulate biomarker data for disease classification
disease_data <- tibble(
  patient_id = paste0("P", str_pad(1:300, 3, pad = "0")),
  # Biomarkers
  biomarker_1 = rnorm(300, 100, 20),
  biomarker_2 = rnorm(300, 50, 10),
  biomarker_3 = rnorm(300, 75, 15),
  biomarker_4 = rnorm(300, 30, 8),
  age = sample(25:80, 300, replace = TRUE),
  sex = sample(c("M", "F"), 300, replace = TRUE)
) %>%
  mutate(
    # Disease status based on biomarkers (with some noise)
    disease_score = 0.5 * biomarker_1 + 0.3 * biomarker_2 -
                    0.2 * biomarker_3 + rnorm(300, 0, 10),
    disease = if_else(disease_score > 60, "diseased", "healthy"),
    disease = factor(disease, levels = c("healthy", "diseased"))
  ) %>%
  select(-disease_score)

# Overview
disease_data %>%
  count(disease)

glimpse(disease_data)
```

## Step 1: Train/Test Split

```{r train-test-split}
# Split data: 75% training, 25% testing
set.seed(123)
data_split <- initial_split(disease_data, prop = 0.75, strata = disease)

# Extract training and testing sets
train_data <- training(data_split)
test_data <- testing(data_split)

# Check distribution
train_data %>% count(disease)
test_data %>% count(disease)
```

## Step 2: Feature Engineering with Recipes

```{r recipe}
# Create preprocessing recipe
disease_recipe <- recipe(disease ~ ., data = train_data) %>%
  # Remove ID column
  update_role(patient_id, new_role = "ID") %>%

  # Normalize numeric predictors
  step_normalize(all_numeric_predictors()) %>%

  # Convert categorical to dummy variables
  step_dummy(all_nominal_predictors()) %>%

  # Remove zero-variance predictors
  step_zv(all_predictors())

disease_recipe

# Prepare the recipe (estimate parameters from training data)
disease_prep <- prep(disease_recipe)

# Apply to training data
train_processed <- bake(disease_prep, new_data = NULL)
head(train_processed)

# Apply to test data
test_processed <- bake(disease_prep, new_data = test_data)
```

## Step 3: Model Specification

### Logistic Regression

```{r logistic-spec}
# Specify logistic regression model
logistic_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

logistic_spec
```

### Random Forest

```{r rf-spec}
# Specify random forest
rf_spec <- rand_forest(trees = 500) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

rf_spec
```

## Step 4: Workflows

Combine recipe and model:

```{r workflow}
# Logistic regression workflow
logistic_wf <- workflow() %>%
  add_recipe(disease_recipe) %>%
  add_model(logistic_spec)

# Random forest workflow
rf_wf <- workflow() %>%
  add_recipe(disease_recipe) %>%
  add_model(rf_spec)
```

## Step 5: Train Models

```{r train-models}
# Train logistic regression
logistic_fit <- logistic_wf %>%
  fit(data = train_data)

# Train random forest
rf_fit <- rf_wf %>%
  fit(data = train_data)
```

## Step 6: Make Predictions

```{r predictions}
# Predict on test data
logistic_pred <- predict(logistic_fit, test_data) %>%
  bind_cols(predict(logistic_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(disease))

rf_pred <- predict(rf_fit, test_data) %>%
  bind_cols(predict(rf_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(disease))

head(logistic_pred)
```

## Step 7: Evaluate Models

```{r evaluate}
# Accuracy
logistic_acc <- accuracy(logistic_pred, truth = disease, estimate = .pred_class)
rf_acc <- accuracy(rf_pred, truth = disease, estimate = .pred_class)

# Combine metrics
metrics <- bind_rows(
  logistic_acc %>% mutate(model = "Logistic Regression"),
  rf_acc %>% mutate(model = "Random Forest")
)

metrics

# Multiple metrics at once
logistic_metrics <- logistic_pred %>%
  metrics(truth = disease, estimate = .pred_class, .pred_diseased)

rf_metrics <- rf_pred %>%
  metrics(truth = disease, estimate = .pred_class, .pred_diseased)

bind_rows(
  logistic_metrics %>% mutate(model = "Logistic"),
  rf_metrics %>% mutate(model = "Random Forest")
)
```

### Confusion Matrix

```{r confusion-matrix}
# Logistic regression
conf_mat(logistic_pred, truth = disease, estimate = .pred_class)

# Random forest
conf_mat(rf_pred, truth = disease, estimate = .pred_class)

# Visualize
conf_mat(rf_pred, truth = disease, estimate = .pred_class) %>%
  autoplot(type = "heatmap")
```

### ROC Curves

```{r roc-curves}
# Calculate ROC curves
logistic_roc <- logistic_pred %>%
  roc_curve(truth = disease, .pred_diseased)

rf_roc <- rf_pred %>%
  roc_curve(truth = disease, .pred_diseased)

# Plot
bind_rows(
  logistic_roc %>% mutate(model = "Logistic"),
  rf_roc %>% mutate(model = "Random Forest")
) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(size = 1.5) +
  geom_abline(linetype = "dashed") +
  labs(title = "ROC Curves",
       x = "1 - Specificity (FPR)",
       y = "Sensitivity (TPR)",
       color = "Model") +
  theme_minimal()

# AUC
logistic_auc <- roc_auc(logistic_pred, truth = disease, .pred_diseased)
rf_auc <- roc_auc(rf_pred, truth = disease, .pred_diseased)

bind_rows(
  logistic_auc %>% mutate(model = "Logistic"),
  rf_auc %>% mutate(model = "Random Forest")
)
```

## Cross-Validation

More robust evaluation using resampling:

```{r cross-validation}
# Create 10-fold cross-validation
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 10, strata = disease)

cv_folds

# Fit models to each fold
logistic_cv <- logistic_wf %>%
  fit_resamples(
    resamples = cv_folds,
    metrics = metric_set(accuracy, roc_auc, sens, spec)
  )

rf_cv <- rf_wf %>%
  fit_resamples(
    resamples = cv_folds,
    metrics = metric_set(accuracy, roc_auc, sens, spec)
  )

# Collect metrics
collect_metrics(logistic_cv)
collect_metrics(rf_cv)

# Compare
bind_rows(
  collect_metrics(logistic_cv) %>% mutate(model = "Logistic"),
  collect_metrics(rf_cv) %>% mutate(model = "Random Forest")
) %>%
  ggplot(aes(x = model, y = mean, fill = model)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err),
                width = 0.2) +
  facet_wrap(~.metric, scales = "free_y") +
  labs(title = "Cross-Validation Performance",
       y = "Mean ± SE") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Hyperparameter Tuning

Optimize model parameters:

```{r tuning}
# Specify model with tunable parameters
rf_tune_spec <- rand_forest(
  mtry = tune(),      # Number of predictors to sample
  trees = 500,
  min_n = tune()      # Minimum node size
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# Create workflow
rf_tune_wf <- workflow() %>%
  add_recipe(disease_recipe) %>%
  add_model(rf_tune_spec)

# Create tuning grid
rf_grid <- grid_regular(
  mtry(range = c(2, 6)),
  min_n(range = c(2, 10)),
  levels = 5
)

rf_grid

# Tune model
set.seed(123)
rf_tune_results <- rf_tune_wf %>%
  tune_grid(
    resamples = cv_folds,
    grid = rf_grid,
    metrics = metric_set(roc_auc, accuracy)
  )

# View results
collect_metrics(rf_tune_results)

# Visualize tuning results
autoplot(rf_tune_results)

# Best parameters
show_best(rf_tune_results, metric = "roc_auc")

# Select best model
best_rf <- select_best(rf_tune_results, metric = "roc_auc")
best_rf
```

### Finalize and Fit Best Model

```{r finalize-model}
# Update workflow with best parameters
final_rf_wf <- rf_tune_wf %>%
  finalize_workflow(best_rf)

# Fit on full training data
final_rf_fit <- final_rf_wf %>%
  fit(data = train_data)

# Evaluate on test data
final_pred <- predict(final_rf_fit, test_data) %>%
  bind_cols(predict(final_rf_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(disease))

# Metrics
final_pred %>%
  metrics(truth = disease, estimate = .pred_class, .pred_diseased)
```

## Feature Importance

Which predictors matter most?

```{r variable-importance}
# Extract fitted model
final_rf_model <- final_rf_fit %>%
  extract_fit_parsnip()

# Variable importance plot
vip(final_rf_model, num_features = 10)

# As tibble
vip_data <- final_rf_model %>%
  vip::vi() %>%
  arrange(desc(Importance))

vip_data

# Custom plot
ggplot(vip_data, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Feature Importance",
       x = "Variable",
       y = "Importance") +
  theme_minimal()
```

## Regression Example: Predicting Gene Expression

```{r regression-example}
# Simulate gene expression data
expression_data <- tibble(
  sample_id = paste0("S", 1:200),
  gene_1 = rnorm(200, 100, 30),
  gene_2 = rnorm(200, 150, 40),
  gene_3 = rnorm(200, 75, 25),
  gene_4 = rnorm(200, 200, 50),
  treatment_dose = runif(200, 0, 10)
) %>%
  mutate(
    # Target gene expression depends on other genes
    target_gene = 50 + 0.5 * gene_1 + 0.3 * gene_2 -
                  0.2 * gene_3 + 0.1 * treatment_dose +
                  rnorm(200, 0, 10)
  )

# Split data
set.seed(123)
expr_split <- initial_split(expression_data, prop = 0.75)
expr_train <- training(expr_split)
expr_test <- testing(expr_split)

# Recipe
expr_recipe <- recipe(target_gene ~ ., data = expr_train) %>%
  update_role(sample_id, new_role = "ID") %>%
  step_normalize(all_predictors())

# Model specs
lm_spec <- linear_reg() %>%
  set_engine("lm")

rf_reg_spec <- rand_forest(trees = 500) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Workflows
lm_wf <- workflow() %>%
  add_recipe(expr_recipe) %>%
  add_model(lm_spec)

rf_reg_wf <- workflow() %>%
  add_recipe(expr_recipe) %>%
  add_model(rf_reg_spec)

# Fit models
lm_fit <- lm_wf %>% fit(expr_train)
rf_reg_fit <- rf_reg_wf %>% fit(expr_train)

# Predictions
lm_pred <- predict(lm_fit, expr_test) %>%
  bind_cols(expr_test %>% select(target_gene))

rf_reg_pred <- predict(rf_reg_fit, expr_test) %>%
  bind_cols(expr_test %>% select(target_gene))

# Metrics for regression
metrics_reg <- metric_set(rmse, rsq, mae)

bind_rows(
  metrics_reg(lm_pred, truth = target_gene, estimate = .pred) %>%
    mutate(model = "Linear Regression"),
  metrics_reg(rf_reg_pred, truth = target_gene, estimate = .pred) %>%
    mutate(model = "Random Forest")
)

# Visualize predictions
bind_rows(
  lm_pred %>% mutate(model = "Linear Regression"),
  rf_reg_pred %>% mutate(model = "Random Forest")
) %>%
  ggplot(aes(x = target_gene, y = .pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~model) +
  labs(title = "Predicted vs Actual Gene Expression",
       x = "Actual Expression",
       y = "Predicted Expression") +
  theme_minimal()
```

## Regularized Regression (LASSO)

Feature selection for high-dimensional data:

```{r lasso}
# LASSO model (L1 regularization)
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")

# Workflow
lasso_wf <- workflow() %>%
  add_recipe(expr_recipe) %>%
  add_model(lasso_spec)

# Create CV folds
set.seed(123)
expr_folds <- vfold_cv(expr_train, v = 5)

# Tune penalty parameter
lambda_grid <- grid_regular(penalty(range = c(-3, 0)), levels = 50)

lasso_tune <- lasso_wf %>%
  tune_grid(
    resamples = expr_folds,
    grid = lambda_grid,
    metrics = metric_set(rmse, rsq)
  )

# Best penalty
autoplot(lasso_tune)

best_penalty <- select_best(lasso_tune, metric = "rmse")
best_penalty

# Finalize and fit
final_lasso <- lasso_wf %>%
  finalize_workflow(best_penalty) %>%
  fit(expr_train)

# Extract coefficients
final_lasso %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(estimate != 0)  # Non-zero coefficients
```

## Complete ML Pipeline Function

```{r ml-pipeline}
# Reusable ML pipeline
run_ml_classification <- function(data, outcome, predictors, model_type = "rf") {

  # Formula
  formula <- as.formula(paste(outcome, "~ ."))

  # Split
  data_split <- initial_split(data, prop = 0.75, strata = all_of(outcome))
  train <- training(data_split)
  test <- testing(data_split)

  # Recipe
  rec <- recipe(formula, data = train) %>%
    step_normalize(all_numeric_predictors()) %>%
    step_dummy(all_nominal_predictors()) %>%
    step_zv(all_predictors())

  # Model
  if (model_type == "rf") {
    mod <- rand_forest(trees = 500) %>%
      set_engine("ranger") %>%
      set_mode("classification")
  } else {
    mod <- logistic_reg() %>%
      set_engine("glm") %>%
      set_mode("classification")
  }

  # Workflow
  wf <- workflow() %>%
    add_recipe(rec) %>%
    add_model(mod)

  # Fit
  fit <- wf %>% fit(train)

  # Predict
  pred <- predict(fit, test) %>%
    bind_cols(predict(fit, test, type = "prob")) %>%
    bind_cols(test %>% select(all_of(outcome)))

  # Metrics
  mets <- pred %>%
    metrics(truth = !!sym(outcome),
            estimate = .pred_class,
            !!sym(paste0(".pred_", levels(data[[outcome]])[2])))

  list(
    workflow = wf,
    fit = fit,
    predictions = pred,
    metrics = mets
  )
}

# Use the pipeline
results <- run_ml_classification(disease_data, "disease",
                                  c("biomarker_1", "biomarker_2",
                                    "biomarker_3", "biomarker_4", "age", "sex"))

results$metrics
```

## Multi-Class Classification

```{r multi-class}
# Simulate data with 3 classes
library(palmerpenguins)

penguins_ml <- penguins %>%
  filter(!is.na(sex), !is.na(body_mass_g)) %>%
  select(species, bill_length_mm, bill_depth_mm,
         flipper_length_mm, body_mass_g, sex)

# Split
set.seed(123)
peng_split <- initial_split(penguins_ml, prop = 0.75, strata = species)
peng_train <- training(peng_split)
peng_test <- testing(peng_split)

# Recipe
peng_recipe <- recipe(species ~ ., data = peng_train) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())

# Random forest for multi-class
peng_rf_spec <- rand_forest(trees = 500) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Workflow and fit
peng_wf <- workflow() %>%
  add_recipe(peng_recipe) %>%
  add_model(peng_rf_spec)

peng_fit <- peng_wf %>% fit(peng_train)

# Predictions
peng_pred <- predict(peng_fit, peng_test) %>%
  bind_cols(predict(peng_fit, peng_test, type = "prob")) %>%
  bind_cols(peng_test %>% select(species))

# Multi-class metrics
peng_pred %>%
  metrics(truth = species, estimate = .pred_class)

# Confusion matrix
conf_mat(peng_pred, truth = species, estimate = .pred_class) %>%
  autoplot(type = "heatmap")
```

## Practice Exercises

```{r exercises, eval=FALSE}
# Exercise 1: Build a random forest model to predict penguin sex
# from body measurements. What's the accuracy? Which features matter most?


# Exercise 2: Create a regression model to predict bill length from
# other measurements. Compare linear regression vs random forest.


# Exercise 3: Tune hyperparameters for a random forest predicting disease
# status. Try different values of mtry and min_n.


# Exercise 4: Build a LASSO model for feature selection on high-dimensional
# data. Which features have non-zero coefficients?


# Exercise 5: Create a complete ML pipeline that:
# - Splits data
# - Preprocesses with a recipe
# - Tunes hyperparameters
# - Evaluates on test set
# - Returns feature importance


```

## Model Comparison Summary

```{r comparison-table, echo=FALSE}
tibble(
  Model = c("Logistic Regression", "Random Forest", "LASSO",
            "Linear Regression", "Poisson Regression"),
  `Use Case` = c("Binary classification", "Classification/Regression",
                 "High-dimensional feature selection",
                 "Continuous outcome", "Count outcome"),
  Pros = c("Interpretable, fast", "Handles interactions, non-linear",
           "Feature selection, prevents overfitting",
           "Simple, interpretable", "Appropriate for counts"),
  Cons = c("Assumes linear relationship", "Black box, slower",
           "Requires tuning", "Assumes linearity", "Requires appropriate distribution")
) %>%
  knitr::kable()
```

## Key Takeaways

- **tidymodels** provides consistent interface for ML in R
- **rsample** for splitting and resampling
- **recipes** for preprocessing and feature engineering
- **parsnip** for model specification
- **workflows** combine recipes and models
- **tune** for hyperparameter optimization
- **yardstick** for evaluation metrics
- Always split data BEFORE any processing
- Use cross-validation for robust evaluation
- Tune hyperparameters systematically
- Check multiple metrics, not just accuracy
- Feature importance reveals what drives predictions
- Different models for different problem types

## Best Practices

1. **Always split first** - prevent data leakage
2. **Stratify splits** - maintain class balance
3. **Use CV** - more robust than single train/test
4. **Tune hyperparameters** - don't use defaults blindly
5. **Multiple metrics** - accuracy isn't everything
6. **Feature importance** - understand your model
7. **Test on held-out data** - final evaluation
8. **Document preprocessing** - recipes are reproducible

## Next Steps

You can now:
- Build predictive models for biological data
- Evaluate model performance properly
- Tune models for optimal performance
- Select important features
- Apply ML to classification and regression problems
- Use tidymodels framework for reproducible ML

---

**Homework:**

1. Choose a biological dataset with a categorical outcome
2. Build 3 different models (logistic, random forest, etc.)
3. Use cross-validation to compare them
4. Tune the best model's hyperparameters
5. Evaluate on test set
6. Create feature importance plot
7. Write interpretation: Which features predict the outcome? How well does the model perform?
