---
title: "Lesson 11: purrr and Functional Programming"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Learning Objectives

By the end of this lesson, you will be able to:

1. Understand functional programming concepts in R
2. Use `map()` and its variants to iterate over data
3. Apply functions to multiple datasets simultaneously
4. Work with list columns in data frames
5. Use `map2()` and `pmap()` for parallel iteration
6. Handle errors gracefully with `safely()` and `possibly()`
7. Apply these skills to real biological data workflows

## Why purrr?

You've used loops before:

```{r traditional-loop, eval=FALSE}
results <- list()
for (i in 1:length(samples)) {
  results[[i]] <- analyze(samples[[i]])
}
```

**purrr makes this cleaner, safer, and more readable!**

## Setup

```{r load-packages}
library(tidyverse)  # includes purrr
library(broom)      # for tidying statistical outputs

set.seed(42)
```

## The map() Family

`map()` applies a function to each element of a list or vector.

### Basic map()

```{r map-basics}
# Simple example
numbers <- list(1, 2, 3, 4, 5)

# Square each number
map(numbers, ~.x^2)

# The ~.x is shorthand for function(x) x^2
# .x refers to each element
```

### map() with Real Data

```{r map-real}
# Multiple samples, calculate mean for each
sample_data <- list(
  control_1 = c(23, 25, 24, 26, 25),
  control_2 = c(22, 24, 23, 25, 24),
  treated_1 = c(45, 47, 46, 48, 47),
  treated_2 = c(44, 46, 45, 47, 46)
)

# Calculate mean of each sample
map(sample_data, mean)

# Get summary statistics
map(sample_data, summary)
```

### map_*() Variants - Type-Stable Output

```{r map-variants}
# map_dbl() returns a numeric vector
map_dbl(sample_data, mean)

# map_chr() returns a character vector
map_chr(sample_data, ~paste("Mean:", round(mean(.x), 2)))

# map_int() returns an integer vector
map_int(sample_data, length)

# map_lgl() returns a logical vector
map_lgl(sample_data, ~mean(.x) > 30)

# map_df() returns a data frame
map_df(sample_data, ~tibble(mean = mean(.x), sd = sd(.x)), .id = "sample")
```

## Practical Example: Reading Multiple Files

Imagine you have data from multiple experiments:

```{r multiple-files-create}
# Create example data files
experiment_data <- list(
  exp1 = tibble(gene = paste0("Gene_", 1:5), expression = rnorm(5, 100, 20)),
  exp2 = tibble(gene = paste0("Gene_", 1:5), expression = rnorm(5, 120, 25)),
  exp3 = tibble(gene = paste0("Gene_", 1:5), expression = rnorm(5, 90, 15))
)

# Save to CSV files
walk2(experiment_data, names(experiment_data),
      ~write_csv(.x, paste0(.y, ".csv")))
```

Read them all at once:

```{r multiple-files-read}
# Get file names
file_names <- list.files(pattern = "exp.*\\.csv")

# Read all files
all_experiments <- map(file_names, read_csv)

# Add experiment ID
all_experiments <- map2(all_experiments, file_names,
                        ~mutate(.x, experiment = .y))

# Combine into one data frame
combined_data <- bind_rows(all_experiments)

combined_data
```

Cleaner approach with names:

```{r multiple-files-clean}
# Read with names
experiments <- set_names(file_names) %>%
  map_df(read_csv, .id = "experiment")

experiments
```

## map2() - Iterate Over Two Inputs

When you need to iterate over two things simultaneously:

```{r map2}
# Two vectors
samples <- c("Sample_A", "Sample_B", "Sample_C")
concentrations <- c(10, 20, 30)

# Combine them
map2_chr(samples, concentrations,
         ~paste(.x, "at", .y, "μg/mL"))

# More complex example: t-tests on paired data
control_groups <- list(
  c(23, 25, 24, 26, 25),
  c(45, 47, 46, 48, 47),
  c(67, 69, 68, 70, 69)
)

treatment_groups <- list(
  c(34, 36, 35, 37, 36),
  c(52, 54, 53, 55, 54),
  c(71, 73, 72, 74, 73)
)

# Perform t-tests
results <- map2(control_groups, treatment_groups, t.test)

# Extract p-values
map_dbl(results, "p.value")
```

## pmap() - Iterate Over Multiple Inputs

When you have more than two inputs:

```{r pmap}
# Data frame with parameters
pcr_conditions <- tibble(
  template = c(10, 20, 30, 40),
  primer = c(0.2, 0.4, 0.2, 0.4),
  cycles = c(30, 30, 35, 35)
)

# Function that uses all three parameters
simulate_pcr <- function(template, primer, cycles) {
  # Simplified PCR simulation
  amplification <- template * (1.9^cycles) * (primer * 5)
  return(amplification)
}

# Apply to all combinations
pcr_conditions %>%
  mutate(yield = pmap_dbl(., simulate_pcr))
```

## Working with Data Frames

Split-apply-combine pattern with map:

```{r split-apply-combine}
# Sample data
library(palmerpenguins)
penguins <- penguins

# Split by species, apply function, combine
penguin_models <- penguins %>%
  filter(!is.na(body_mass_g), !is.na(flipper_length_mm)) %>%
  split(.$species) %>%
  map(~lm(body_mass_g ~ flipper_length_mm, data = .x))

# Extract R-squared
penguin_models %>%
  map(summary) %>%
  map_dbl("r.squared")

# Extract coefficients nicely with broom
penguin_models %>%
  map_df(tidy, .id = "species")
```

## Real Example: Multiple Statistical Tests

```{r multiple-tests}
# Gene expression data
gene_expression <- tibble(
  gene = paste0("Gene_", 1:20),
  control_1 = rnorm(20, 100, 15),
  control_2 = rnorm(20, 100, 15),
  control_3 = rnorm(20, 100, 15),
  treated_1 = rnorm(20, 130, 20),
  treated_2 = rnorm(20, 130, 20),
  treated_3 = rnorm(20, 130, 20)
)

# Reshape for testing
gene_long <- gene_expression %>%
  pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
  mutate(group = if_else(str_detect(sample, "control"), "control", "treated"))

# Perform t-test for each gene
test_results <- gene_long %>%
  split(.$gene) %>%
  map(~t.test(expression ~ group, data = .x)) %>%
  map_df(tidy, .id = "gene")

# Adjust p-values
test_results %>%
  mutate(p_adjusted = p.adjust(p.value, method = "BH")) %>%
  arrange(p_adjusted)
```

## Error Handling with safely() and possibly()

Not all operations succeed. Handle errors gracefully:

```{r error-handling}
# Some operations might fail
data_list <- list(
  good1 = c(1, 2, 3, 4, 5),
  bad = c(1, 2, NA, 4, 5),  # Has NA
  good2 = c(6, 7, 8, 9, 10)
)

# This would fail on 'bad' if we required no NAs
calculate_stats <- function(x) {
  if (any(is.na(x))) stop("NA values present")
  mean(x)
}

# Using safely() - returns list with result and error
safe_calculate <- safely(calculate_stats)

results <- map(data_list, safe_calculate)

# Extract successful results
map(results, "result")

# Extract errors
map(results, "error")

# Get only successful results
successful <- results %>%
  keep(~is.null(.x$error)) %>%
  map_dbl("result")

successful
```

### possibly() - Provide Default Values

```{r possibly}
# possibly() returns a default value on error
safe_mean <- possibly(calculate_stats, otherwise = NA_real_)

# Now all results succeed
map_dbl(data_list, safe_mean)
```

## Nested Data and List Columns

Store complex data in tibbles:

```{r nested-data}
# Create nested data
growth_data <- tibble(
  strain = c("WT", "mutant_A", "mutant_B"),
  timepoints = list(
    c(0, 2, 4, 6, 8),
    c(0, 2, 4, 6, 8),
    c(0, 2, 4, 6, 8)
  ),
  od600 = list(
    c(0.05, 0.3, 0.8, 1.2, 1.5),
    c(0.05, 0.2, 0.5, 0.8, 1.0),
    c(0.05, 0.35, 0.9, 1.3, 1.6)
  )
)

growth_data

# Calculate growth rate for each strain
growth_data %>%
  mutate(
    max_od = map_dbl(od600, max),
    doubling_time = map2_dbl(timepoints, od600, ~{
      # Simplified calculation
      log_phase <- 2:4  # assuming log phase
      time_diff <- .x[log_phase]
      od_diff <- log2(.y[log_phase])
      1 / coef(lm(od_diff ~ time_diff))[2]
    })
  )
```

## Practical Workflow: Analyzing Multiple Plates

```{r plate-analysis}
# Simulate 96-well plate data for 3 plates
simulate_plate <- function(plate_id) {
  tibble(
    plate = plate_id,
    well = paste0(rep(LETTERS[1:8], each = 12), rep(1:12, 8)),
    row = rep(LETTERS[1:8], each = 12),
    col = rep(1:12, 8),
    absorbance = rnorm(96, mean = 0.5, sd = 0.1) +
                 ifelse(row %in% c("A", "H"), -0.2, 0)  # Edge effects
  )
}

# Generate data for 3 plates
plates <- tibble(plate_id = paste0("Plate_", 1:3)) %>%
  mutate(data = map(plate_id, simulate_plate))

plates

# Unnest to see all data
plates %>%
  unnest(data)

# Calculate stats for each plate
plate_summary <- plates %>%
  mutate(
    mean_abs = map_dbl(data, ~mean(.x$absorbance)),
    median_abs = map_dbl(data, ~median(.x$absorbance)),
    cv = map_dbl(data, ~sd(.x$absorbance) / mean(.x$absorbance) * 100),
    edge_effect = map_lgl(data, ~{
      edge_wells <- filter(.x, row %in% c("A", "H"))
      inner_wells <- filter(.x, !row %in% c("A", "H"))
      t.test(edge_wells$absorbance, inner_wells$absorbance)$p.value < 0.05
    })
  )

plate_summary
```

## Advanced: Custom Functions with purrr

```{r custom-functions}
# Analyze growth curves
analyze_growth <- function(time, od) {
  # Fit exponential growth model
  model <- nls(od ~ a * exp(b * time),
               start = list(a = min(od), b = 0.1),
               control = nls.control(maxiter = 100))

  # Extract parameters
  tibble(
    initial_od = coef(model)[1],
    growth_rate = coef(model)[2],
    r_squared = cor(od, predict(model))^2
  )
}

# Safe version
safe_analyze <- possibly(analyze_growth, otherwise = tibble(
  initial_od = NA_real_,
  growth_rate = NA_real_,
  r_squared = NA_real_
))

# Apply to growth data
growth_params <- growth_data %>%
  mutate(params = map2(timepoints, od600, safe_analyze)) %>%
  unnest(params)

growth_params
```

## Combining Everything: High-Throughput qPCR Analysis

```{r qpcr-workflow}
# Simulate qPCR data from multiple runs
generate_qpcr_data <- function(run_id) {
  tibble(
    run = run_id,
    gene = rep(c("actb", "gapdh", "target1", "target2"), each = 3),
    replicate = rep(1:3, 4),
    ct = c(
      rnorm(3, 18, 0.5),    # actb
      rnorm(3, 19, 0.5),    # gapdh
      rnorm(3, 24, 0.8),    # target1
      rnorm(3, 26, 0.8)     # target2
    )
  )
}

# Generate data for 5 runs
qpcr_runs <- tibble(run_id = paste0("Run_", 1:5)) %>%
  mutate(data = map(run_id, generate_qpcr_data))

# Quality control function
qc_check <- function(data) {
  data %>%
    group_by(gene) %>%
    summarise(
      mean_ct = mean(ct),
      sd_ct = sd(ct),
      cv = sd_ct / mean_ct * 100,
      pass_qc = cv < 5  # CV threshold
    )
}

# Apply QC to each run
qpcr_qc <- qpcr_runs %>%
  mutate(qc_results = map(data, qc_check))

# View QC results
qpcr_qc %>%
  select(run_id, qc_results) %>%
  unnest(qc_results)

# Calculate delta-delta Ct for passing runs
calculate_ddct <- function(data) {
  data %>%
    group_by(gene) %>%
    summarise(mean_ct = mean(ct)) %>%
    mutate(
      reference_ct = mean_ct[gene == "actb"],
      delta_ct = mean_ct - reference_ct,
      delta_delta_ct = delta_ct - delta_ct[gene == "target1"],
      fold_change = 2^(-delta_delta_ct)
    ) %>%
    filter(gene %in% c("target1", "target2"))
}

# Apply to all runs
qpcr_results <- qpcr_runs %>%
  mutate(ddct = map(data, calculate_ddct))

# Final results
qpcr_results %>%
  select(run_id, ddct) %>%
  unnest(ddct)
```

## walk() - For Side Effects

When you want to DO something but not return results:

```{r walk}
# Save each plate's data
plates %>%
  pwalk(function(plate_id, data) {
    filename <- paste0(plate_id, "_data.csv")
    write_csv(data, filename)
  })

# Create plots for each strain
walk(unique(growth_data$strain), function(s) {
  plot_data <- growth_data %>%
    filter(strain == s) %>%
    unnest(c(timepoints, od600))

  p <- ggplot(plot_data, aes(x = timepoints, y = od600)) +
    geom_line(size = 1) +
    geom_point(size = 2) +
    labs(title = paste("Growth curve:", s),
         x = "Time (hours)", y = "OD600") +
    theme_minimal()

  ggsave(paste0("growth_", s, ".png"), p, width = 6, height = 4)
})
```

## Practice Exercises

```{r exercises, eval=FALSE}
# Exercise 1: You have 10 samples. Calculate mean, median, and sd for each.
samples <- map(1:10, ~rnorm(20, mean = 100 + .x * 5, sd = 10))


# Exercise 2: Read multiple CSV files and filter each to keep only rows
# where a specific column > threshold


# Exercise 3: Split the penguins data by island, fit a linear model
# (body_mass ~ flipper_length) for each, extract R-squared values


# Exercise 4: Create a nested tibble with different growth conditions,
# then calculate max growth rate for each


# Exercise 5: Use map2() to perform multiple t-tests between paired
# control and treatment groups, then extract p-values and adjust for
# multiple testing


```

## Common Patterns Cheat Sheet

```{r cheat-sheet, eval=FALSE}
# Read multiple files
files %>% map_df(read_csv, .id = "file")

# Apply function to groups
data %>% split(.$group) %>% map(analyze)

# Safe operations
data %>% map(possibly(risky_function, otherwise = NA))

# Extract element from list
results %>% map_dbl("p.value")

# Parallel iteration
map2(x, y, ~some_function(.x, .y))

# Multiple inputs
pmap(list(x, y, z), ~some_function(..1, ..2, ..3))

# Side effects only
walk(data, save_to_file)

# Keep only successful results
results %>% keep(~is.null(.x$error))
```

## When to Use purrr vs. Loops

**Use purrr when:**
- Applying same function to multiple inputs
- Working with list columns
- Need type-stable output
- Want cleaner, more readable code
- Need error handling

**Use loops when:**
- Complex iteration logic
- Need to modify in place
- Breaking out early based on conditions
- Learning/teaching iteration basics

## Key Takeaways

- `map()` applies a function to each element of a list/vector
- `map_*()` variants return specific types (dbl, chr, df, etc.)
- `map2()` and `pmap()` handle multiple parallel inputs
- `safely()` and `possibly()` handle errors gracefully
- List columns in tibbles enable complex data structures
- `walk()` is for side effects (saving files, creating plots)
- purrr makes iteration cleaner and safer than loops
- Perfect for high-throughput biological data analysis

## Next Steps

In Lesson 12, we'll learn stringr for text manipulation - essential for working with gene names, sequences, and sample IDs!

---

**Homework:**

1. Take a dataset with multiple groups (like penguins by species)
2. Split it using `split()`
3. Fit a model to each group using `map()`
4. Extract statistics with `map_df()` and broom's `tidy()`
5. Create a plot for each group using `walk()`

Clean up:

```{r cleanup, include=FALSE}
# Remove created files
file.remove(list.files(pattern = "exp.*\\.csv"))
file.remove(list.files(pattern = "Plate.*\\.csv"))
file.remove(list.files(pattern = "growth_.*\\.png"))
```
