---
title: "Lesson 4: Grouping and Summarizing Data"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Learning Objectives

By the end of this lesson, you will be able to:

1. Use `group_by()` to create groups in your data
2. Use `summarise()` to calculate summary statistics
3. Combine grouping and summarizing for powerful analyses
4. Count observations with `count()` and `n()`
5. Calculate multiple statistics at once
6. Apply these skills to experimental data analysis

## Setup

```{r load-packages}
library(tidyverse)
library(palmerpenguins)

penguins <- penguins
```

## Why Group and Summarize?

In biological research, we constantly compare groups:

- Control vs Treatment
- Male vs Female
- Species A vs Species B vs Species C
- Time points: Day 0, Day 1, Day 2...

Grouping and summarizing lets us calculate statistics for each group automatically!

## summarise(): Calculate Summary Statistics

`summarise()` (or `summarize()`) reduces your data to summary statistics.

### Basic Summaries

```{r summarise-basic}
# Overall mean body mass
penguins %>%
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE))

# Multiple statistics at once
penguins %>%
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    min_mass = min(body_mass_g, na.rm = TRUE),
    max_mass = max(body_mass_g, na.rm = TRUE),
    n_penguins = n()  # count of observations
  )
```

## group_by(): The Real Power

`group_by()` tells R to perform operations separately for each group.

### Grouping by One Variable

```{r group-one}
# Mean body mass by species
penguins %>%
  group_by(species) %>%
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    n = n()
  )
```

This is much more useful than the overall mean! We can see Gentoo penguins are much larger.

### More Statistics by Group

```{r group-stats}
# Comprehensive summary by species
species_summary <- penguins %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
    sd_flipper = sd(flipper_length_mm, na.rm = TRUE)
  )

species_summary
```

### Grouping by Multiple Variables

```{r group-multiple}
# Mean body mass by species AND sex
penguins %>%
  group_by(species, sex) %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE)
  )
```

Notice males are heavier than females in all species!

### Grouping by Three Variables

```{r group-three}
# Mean body mass by species, sex, AND island
penguins %>%
  group_by(species, island, sex) %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE)
  ) %>%
  arrange(species, island, sex)
```

## Common Summary Functions

Here are functions you'll use frequently:

```{r summary-functions}
penguins %>%
  group_by(species) %>%
  summarise(
    # Central tendency
    mean_bill = mean(bill_length_mm, na.rm = TRUE),
    median_bill = median(bill_length_mm, na.rm = TRUE),

    # Spread
    sd_bill = sd(bill_length_mm, na.rm = TRUE),
    var_bill = var(bill_length_mm, na.rm = TRUE),

    # Range
    min_bill = min(bill_length_mm, na.rm = TRUE),
    max_bill = max(bill_length_mm, na.rm = TRUE),

    # Count
    n_obs = n(),
    n_missing = sum(is.na(bill_length_mm))
  )
```

## count(): A Shortcut for Counting

`count()` is a quick way to count observations in groups:

```{r count}
# How many penguins per species?
penguins %>%
  count(species)

# By species and island
penguins %>%
  count(species, island)

# Sort by count
penguins %>%
  count(species, sort = TRUE)
```

This is equivalent to:

```{r count-equivalent}
penguins %>%
  group_by(species) %>%
  summarise(n = n())
```

## Calculating Proportions

```{r proportions}
# What proportion of penguins are each species?
penguins %>%
  count(species) %>%
  mutate(proportion = n / sum(n))

# As percentages
penguins %>%
  count(species) %>%
  mutate(
    proportion = n / sum(n),
    percent = round(proportion * 100, 1)
  )
```

## Filtering After Summarizing

You can filter summary results:

```{r filter-summary}
# Only show species with mean body mass > 4500g
penguins %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE)
  ) %>%
  filter(mean_mass > 4500)
```

## Combining with Other dplyr Verbs

The real power comes from combining operations:

### Example 1: Filter, Group, Summarize

```{r combine-1}
# Summary statistics for only 2009 data
penguins %>%
  filter(year == 2009) %>%
  group_by(species, sex) %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    se_mass = sd(body_mass_g, na.rm = TRUE) / sqrt(n())  # standard error
  )
```

### Example 2: Group, Summarize, Arrange

```{r combine-2}
# Which species-island combo has the highest mean body mass?
penguins %>%
  group_by(species, island) %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_mass))
```

### Example 3: Complex Pipeline

```{r combine-3}
# Find islands where female penguins average > 4000g
penguins %>%
  filter(sex == "female", !is.na(body_mass_g)) %>%
  group_by(island) %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g),
    sd_mass = sd(body_mass_g)
  ) %>%
  filter(mean_mass > 4000) %>%
  arrange(desc(mean_mass))
```

## Real-World Biology Example: Cell Growth

```{r biology-example}
# Simulated bacterial growth experiment
growth_data <- tibble(
  strain = rep(c("WT", "mutant_A", "mutant_B"), each = 12),
  replicate = rep(1:3, times = 12),
  timepoint_hr = rep(c(0, 2, 4, 6), each = 3, times = 3),
  od600 = c(
    # WT
    0.05, 0.05, 0.05,  # 0hr
    0.3, 0.32, 0.28,   # 2hr
    0.85, 0.90, 0.82,  # 4hr
    1.2, 1.25, 1.18,   # 6hr
    # mutant_A (slower growth)
    0.05, 0.05, 0.05,
    0.15, 0.18, 0.16,
    0.45, 0.48, 0.43,
    0.72, 0.75, 0.70,
    # mutant_B (similar to WT)
    0.05, 0.05, 0.05,
    0.28, 0.30, 0.29,
    0.80, 0.85, 0.82,
    1.15, 1.20, 1.18
  )
)

# Calculate mean and SE for each strain at each timepoint
growth_summary <- growth_data %>%
  group_by(strain, timepoint_hr) %>%
  summarise(
    n = n(),
    mean_od600 = mean(od600),
    sd_od600 = sd(od600),
    se_od600 = sd_od600 / sqrt(n)
  )

growth_summary

# Which strain grows fastest at 6 hours?
growth_data %>%
  filter(timepoint_hr == 6) %>%
  group_by(strain) %>%
  summarise(mean_od600 = mean(od600)) %>%
  arrange(desc(mean_od600))
```

## Visualizing Summarized Data

Summarized data is perfect for plotting:

```{r plot-summary}
# Plot species means with error bars
penguins %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    se_mass = sd(body_mass_g, na.rm = TRUE) / sqrt(n)
  ) %>%
  ggplot(aes(x = species, y = mean_mass)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_mass - se_mass,
                    ymax = mean_mass + se_mass),
                width = 0.2) +
  labs(title = "Mean Body Mass by Species",
       x = "Species",
       y = "Body Mass (g) ± SE") +
  theme_minimal()
```

## Practice Exercises

```{r exercises, eval=FALSE}
# Exercise 1: Calculate mean and SD of flipper length for each island


# Exercise 2: How many penguins of each sex were observed each year?


# Exercise 3: Which species-sex combination has the longest average bill?


# Exercise 4: For each species, what percentage of observations are from each island?


# Exercise 5: Calculate the coefficient of variation (CV = sd/mean * 100)
#             of body mass for each species. Which is most variable?


```

## Advanced: Grouping Without Summarizing

Sometimes you want to calculate within groups but keep all rows:

```{r group-mutate}
# Calculate z-scores within each species
penguins %>%
  group_by(species) %>%
  mutate(
    species_mean_mass = mean(body_mass_g, na.rm = TRUE),
    species_sd_mass = sd(body_mass_g, na.rm = TRUE),
    z_score = (body_mass_g - species_mean_mass) / species_sd_mass
  ) %>%
  select(species, body_mass_g, z_score) %>%
  head(10)
```

Don't forget to `ungroup()` when done:

```{r ungroup}
penguins_grouped <- penguins %>%
  group_by(species)

# Check grouping
groups(penguins_grouped)

# Remove grouping
penguins_ungrouped <- penguins_grouped %>%
  ungroup()

groups(penguins_ungrouped)
```

## Common Pitfalls

### 1. Forgetting na.rm = TRUE

```{r pitfall-1}
# This returns NA because there are missing values
penguins %>%
  group_by(species) %>%
  summarise(mean_mass = mean(body_mass_g))

# This works
penguins %>%
  group_by(species) %>%
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE))
```

### 2. Grouping Variables in Output

```{r pitfall-2}
# Grouping variables are automatically included
penguins %>%
  group_by(species) %>%
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE))
# species column is automatically there!
```

## Key Takeaways

- `group_by()` creates groups for subsequent operations
- `summarise()` calculates summary statistics
- Together they're powerful for comparing groups
- Always use `na.rm = TRUE` with statistical functions
- `n()` counts observations in each group
- `count()` is a shortcut for `group_by() %>% summarise(n = n())`
- You can group by multiple variables
- Combine with other dplyr verbs for complex analyses

## Next Steps

In Lesson 5, we'll dive deep into data visualization with ggplot2, learning to create publication-quality figures!

---

**Homework:** Analyze the penguins data to answer:

1. What is the mean bill length for each species-sex combination?
2. Which year had the most observations?
3. Calculate mean body mass by species and island, then create a bar plot
4. For each species, what's the difference between male and female mean body mass?
