---
title: "Lesson 18: Reproducible Research"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
params:
  species: "Adelie"
  min_sample_size: 30
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Learning Objectives

By the end of this lesson, you will be able to:

1. Create and use R Projects for organization
2. Structure your analysis directories effectively
3. Use the here package for robust file paths
4. Manage package dependencies with renv
5. Create parameterized RMarkdown reports
6. Use RMarkdown templates
7. Document your analysis environment
8. Apply reproducibility best practices
9. Share your work effectively

## Why Reproducibility Matters

Reproducibility in science means:
- **You** can re-run your analysis in 6 months
- **Collaborators** can understand and extend your work
- **Reviewers** can verify your results
- **The field** can build on your findings

### The Reproducibility Crisis

Common problems:
- "It works on my machine"
- Lost code or data
- Unclear analysis steps
- Different package versions
- Hard-coded file paths
- Manual data manipulation

**We can fix all of these!**

## Setup

```{r load-packages}
library(tidyverse)
library(here)          # Robust file paths
# library(renv)        # Package management (we'll discuss)
library(palmerpenguins)
```

## R Projects: Your Foundation

### What is an R Project?

An `.Rproj` file that:
- Sets working directory automatically
- Keeps work isolated
- Makes paths relative
- Integrates with version control

### Creating a Project

**In RStudio:**
1. File → New Project
2. Choose: New Directory / Existing Directory / Version Control
3. Name your project (e.g., "RNA-seq-analysis")
4. Create Project

**What this does:**
- Creates `project-name.Rproj` file
- Sets working directory to project root
- Isolates your R session

### Project Benefits

```{r project-benefits, eval=FALSE}
# Without project:
setwd("/Users/jane/Documents/Research/Project1/Analysis")
data <- read_csv("/Users/jane/Documents/Research/Project1/Data/samples.csv")
# Breaks on different computer!

# With project:
data <- read_csv("data/samples.csv")
# Works anywhere! (as long as structure is maintained)
```

### Project Settings

**Recommended settings** (Tools → Project Options):

- **Restore .RData:** No (start fresh each time)
- **Save workspace:** No (explicit saving only)
- **Always save history:** Your choice

Why? Fresh starts prevent hidden dependencies!

## Directory Structure

### Best Practice Structure

```
my-analysis/
├── my-analysis.Rproj
├── README.md                 # Project description
├── data/
│   ├── raw/                  # Original, never modified
│   │   └── samples.csv
│   └── processed/            # Cleaned data
│       └── samples_clean.csv
├── scripts/
│   ├── 01_data_cleaning.R
│   ├── 02_analysis.R
│   └── 03_visualization.R
├── reports/
│   ├── analysis_report.Rmd
│   └── analysis_report.html
├── figures/
│   ├── figure1.png
│   └── figure2.png
├── results/
│   ├── model_output.csv
│   └── statistics.txt
├── functions/
│   └── custom_functions.R
└── renv/                     # Package management
    └── renv.lock
```

### Creating Directory Structure

```{r create-structure, eval=FALSE}
# Create all directories at once
dir.create("data/raw", recursive = TRUE)
dir.create("data/processed", recursive = TRUE)
dir.create("scripts")
dir.create("reports")
dir.create("figures")
dir.create("results")
dir.create("functions")
```

## The here Package

Solves the "working directory" problem!

### The Problem

```{r path-problem, eval=FALSE}
# Different working directories cause issues
getwd()  # Where am I?

# These might fail:
read_csv("data/samples.csv")           # If not in project root
read_csv("../data/samples.csv")        # Fragile
read_csv("~/Research/Project/data/samples.csv")  # Not portable
```

### The Solution: here()

```{r here-solution}
library(here)

# Where is project root?
here()

# Build paths relative to project root
here("data", "raw", "samples.csv")
here("figures", "plot1.png")
here("scripts", "analysis.R")

# Works from anywhere in project!
```

### Using here() in Practice

```{r here-practice, eval=FALSE}
# Reading data
samples <- read_csv(here("data", "raw", "samples.csv"))

# Saving results
write_csv(results, here("results", "summary_stats.csv"))

# Saving figures
ggsave(here("figures", "growth_curve.png"), plot = p, width = 8, height = 6)

# Sourcing scripts
source(here("functions", "custom_functions.R"))
```

### Example: Complete Analysis Script

```{r complete-script, eval=FALSE}
# analysis.R - Always works from project root!

library(tidyverse)
library(here)

# Read data
raw_data <- read_csv(here("data", "raw", "expression_data.csv"))

# Process
clean_data <- raw_data %>%
  filter(!is.na(expression)) %>%
  mutate(log_expression = log2(expression + 1))

# Save processed data
write_csv(clean_data, here("data", "processed", "expression_clean.csv"))

# Analyze
results <- clean_data %>%
  group_by(gene) %>%
  summarise(mean_expr = mean(log_expression))

# Save results
write_csv(results, here("results", "gene_summary.csv"))

# Plot
p <- ggplot(results, aes(x = mean_expr)) +
  geom_histogram()

ggsave(here("figures", "expression_distribution.png"), p)
```

## Package Management with renv

### The Problem

```{r package-problem, eval=FALSE}
# Your code today:
library(dplyr)  # version 1.1.0

# Your code in 6 months (dplyr updated to 2.0.0):
library(dplyr)  # version 2.0.0 - breaking changes!
# Code breaks! 😱
```

### The Solution: renv

renv creates a **project-local** package library.

```{r renv-init, eval=FALSE}
# Install renv (once)
install.packages("renv")

# Initialize in your project
renv::init()

# What this does:
# - Creates renv/ directory
# - Creates renv.lock file (package versions)
# - Creates .Rprofile (activates renv on startup)
```

### Basic renv Workflow

```{r renv-workflow, eval=FALSE}
# 1. Start project
renv::activate()

# 2. Install packages (creates project library)
install.packages("tidyverse")
install.packages("ggplot2")

# 3. Work on your analysis...

# 4. Snapshot package versions
renv::snapshot()
# Creates/updates renv.lock with exact versions

# 5. Share project (including renv.lock)
# Collaborator runs:
renv::restore()
# Installs exact same package versions!
```

### Checking Package Status

```{r renv-status, eval=FALSE}
# See what's out of sync
renv::status()

# Update a package
install.packages("dplyr")
renv::snapshot()  # Record new version

# Remove unused packages
renv::clean()
```

### renv.lock File

```json
{
  "R": {
    "Version": "4.3.0"
  },
  "Packages": {
    "dplyr": {
      "Package": "dplyr",
      "Version": "1.1.0",
      "Source": "Repository",
      "Repository": "CRAN"
    }
  }
}
```

This file ensures **exact reproducibility**!

## Parameterized Reports

Make reports flexible and reusable!

### Adding Parameters

In YAML header (top of this document):

```yaml
---
title: "Analysis Report"
params:
  species: "Adelie"
  min_sample_size: 30
---
```

### Using Parameters

```{r use-params}
# Access with params$
species_to_analyze <- params$species
min_n <- params$min_sample_size

cat("Analyzing species:", species_to_analyze, "\n")
cat("Minimum sample size:", min_n, "\n")
```

### Analysis with Parameters

```{r param-analysis}
# Filter based on parameters
selected_data <- penguins %>%
  filter(species == params$species, !is.na(body_mass_g))

# Check sample size
if (nrow(selected_data) < params$min_sample_size) {
  stop("Not enough samples! Need at least ", params$min_sample_size)
}

# Summarize
summary_stats <- selected_data %>%
  summarise(
    n = n(),
    mean_mass = mean(body_mass_g),
    sd_mass = sd(body_mass_g)
  )

summary_stats
```

### Rendering with Different Parameters

```{r render-params, eval=FALSE}
# From R console:
rmarkdown::render(
  "analysis_report.Rmd",
  params = list(species = "Chinstrap", min_sample_size = 20),
  output_file = "chinstrap_report.html"
)

# Create reports for all species
for (sp in c("Adelie", "Chinstrap", "Gentoo")) {
  rmarkdown::render(
    "analysis_report.Rmd",
    params = list(species = sp, min_sample_size = 30),
    output_file = paste0(sp, "_report.html")
  )
}
```

### Interactive Parameters

```{r interactive-params, eval=FALSE}
# Click "Knit with Parameters" button in RStudio
# Or:
rmarkdown::render("report.Rmd", params = "ask")
```

## Advanced RMarkdown

### Child Documents

Break large reports into modules:

```{r child-doc, eval=FALSE}
# In main document:
```

```{r child='sections/data-cleaning.Rmd'}
```

```{r child='sections/analysis.Rmd'}
```

```{r child='sections/visualization.Rmd'}
```

\`\`\`

### Conditional Content

```{r conditional-content}
# Only show detailed diagnostics if requested
show_diagnostics <- FALSE  # Could be a parameter

if (show_diagnostics) {
  cat("## Detailed Diagnostics\n")
  # Complex diagnostic code here
}
```

### Output Formats

```{r output-formats, eval=FALSE}
# YAML header options:

# HTML with table of contents
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide

# PDF
output: pdf_document

# Word
output: word_document

# Multiple formats
output:
  html_document: default
  pdf_document: default
```

## Session Information

**Always include at end of analysis!**

```{r session-info}
# Document your environment
sessionInfo()
```

Better formatted version:

```{r session-info-better}
library(sessioninfo)
session_info()
```

## README Files

Every project needs a README!

### Example README.md

```markdown
# RNA-seq Analysis: Disease vs Control

## Overview
Analysis of RNA-seq data comparing diseased and control samples.

## Structure
- `data/raw/`: Original count files
- `data/processed/`: Normalized counts
- `scripts/`: Analysis scripts (run in order)
- `results/`: Differential expression tables
- `figures/`: Publication-quality figures

## Requirements
- R >= 4.3.0
- See renv.lock for package versions

## Setup
1. Open RNA-seq-analysis.Rproj
2. Run `renv::restore()` to install packages
3. Run scripts in order:
   - 01_data_cleaning.R
   - 02_normalization.R
   - 03_differential_expression.R

## Data
Raw data available at: [DOI/URL]

## Contact
Your Name (email@university.edu)
```

## Code Organization Tips

### Script Headers

```{r script-header, eval=FALSE}
# ============================================================================
# Analysis of Tumor Growth Data
#
# Author: Your Name
# Date: 2024-03-15
# Purpose: Compare tumor volumes between treatment groups
#
# Input: data/raw/tumor_measurements.csv
# Output: results/tumor_stats.csv, figures/growth_curves.png
# ============================================================================

# Load packages ----
library(tidyverse)
library(here)

# Set parameters ----
SIGNIFICANCE_THRESHOLD <- 0.05
MIN_TUMOR_SIZE <- 50  # mm³

# Read data ----
# ...
```

### Section Markers

```{r section-markers, eval=FALSE}
# Data Loading ----

# Data Cleaning ----

# Exploratory Analysis ----

# Statistical Testing ----

# Visualization ----
```

RStudio recognizes `----` and creates code outline!

### Custom Functions

```{r custom-functions-example, eval=FALSE}
# functions/stats_helpers.R

#' Calculate fold change with pseudocount
#'
#' @param treated Vector of treated values
#' @param control Vector of control values
#' @param pseudocount Small value to avoid log(0)
#' @return Numeric vector of fold changes
calculate_fold_change <- function(treated, control, pseudocount = 1) {
  log2((treated + pseudocount) / (control + pseudocount))
}

# In main script:
source(here("functions", "stats_helpers.R"))
fc <- calculate_fold_change(treated_expr, control_expr)
```

## Reproducibility Checklist

### Before Sharing Analysis

- [ ] Code runs from top to bottom without errors
- [ ] All file paths use `here()` or are relative
- [ ] Required packages documented (renv.lock or list)
- [ ] R version documented (session_info())
- [ ] Data sources documented (README)
- [ ] Random seeds set (`set.seed()`)
- [ ] No `install.packages()` in scripts
- [ ] No `setwd()` or absolute paths
- [ ] README explains project structure
- [ ] Sensitive data removed/anonymized

### Documentation to Include

```{r documentation-template, eval=FALSE}
# At end of analysis script or RMarkdown:

# Session Information ----
sessioninfo::session_info()

# File Information ----
cat("Script:", here::here(), "\n")
cat("Date:", Sys.Date(), "\n")
cat("Git commit:", system("git rev-parse HEAD", intern = TRUE), "\n")

# Data provenance ----
cat("Data source: GSE12345 downloaded 2024-03-15\n")
cat("Sample size:", nrow(data), "\n")
```

## Version Control Integration

### .gitignore for R Projects

```bash
# .gitignore

# R specific
.Rproj.user
.Rhistory
.RData
.Ruserdata

# Data (usually don't commit large data files)
data/raw/*.csv
data/raw/*.txt
*.fastq.gz

# Output
*.html
*.pdf
figures/*.png

# renv (commit renv.lock, but not library)
renv/library/
renv/staging/

# Keep these:
!renv.lock
!renv/activate.R
```

## Real-World Example: Complete Project

```{r complete-project, eval=FALSE}
# Project: microbiome-analysis.Rproj

# ── 01_setup.R ──
library(tidyverse)
library(here)

# Initialize renv (once)
renv::init()

# Install packages
install.packages(c("tidyverse", "vegan", "phyloseq"))
renv::snapshot()

# ── 02_import_data.R ──
library(tidyverse)
library(here)

# Read OTU table
otu <- read_tsv(here("data", "raw", "otu_table.txt"))

# Read metadata
metadata <- read_csv(here("data", "raw", "sample_metadata.csv"))

# Save processed
save(otu, metadata, file = here("data", "processed", "data.RData"))

# ── 03_analysis.Rmd ──
---
title: "Microbiome Analysis"
output:
  html_document:
    toc: true
params:
  min_reads: 1000
  treatment: "antibiotic"
---

## Load data
load(here::here("data", "processed", "data.RData"))

## Filter samples
filtered <- otu %>%
  filter(total_reads > params$min_reads)

## Analysis...

## Session Info
sessioninfo::session_info()
```

## Practice Exercises

```{r exercises, eval=FALSE}
# Exercise 1: Create a new R Project with proper directory structure
# Include data/, scripts/, results/, and figures/ folders


# Exercise 2: Write a script that uses here() to:
# - Read data from data/raw/
# - Save processed data to data/processed/
# - Save a figure to figures/


# Exercise 3: Create a parameterized report that takes:
# - dataset name
# - significance threshold
# - output format
# Run it with 3 different parameter sets


# Exercise 4: Set up renv in a project, install packages,
# create snapshot, remove a package, restore from snapshot


# Exercise 5: Create a complete analysis with:
# - Proper directory structure
# - README.md
# - Session info at end
# - All paths using here()
```

## Common Mistakes

1. **Absolute paths**
   ```{r mistake-1, eval=FALSE}
   # Bad
   read_csv("/Users/jane/data.csv")

   # Good
   read_csv(here("data", "raw", "data.csv"))
   ```

2. **Saving workspace**
   ```{r mistake-2, eval=FALSE}
   # Bad - hidden dependencies
   save.image()

   # Good - explicit
   save(important_object, file = here("results", "results.RData"))
   ```

3. **Installing in scripts**
   ```{r mistake-3, eval=FALSE}
   # Bad - runs every time
   install.packages("tidyverse")
   library(tidyverse)

   # Good - separate setup
   # (install once, then just library())
   library(tidyverse)
   ```

## Key Takeaways

- **R Projects** keep work organized and portable
- **here()** makes file paths robust
- **renv** ensures package version reproducibility
- **Directory structure** aids understanding
- **Parameters** make reports flexible
- **Documentation** includes README and session info
- **Version control** tracks changes over time
- **Reproducibility** means anyone (including future you) can run it

## Resources

- [R for Reproducible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/)
- [renv documentation](https://rstudio.github.io/renv/)
- [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
- [Project-oriented workflow](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/)

---

**Homework:**

1. Convert an existing analysis to an R Project
2. Reorganize files into proper directory structure
3. Replace all absolute paths with here()
4. Add renv for package management
5. Create a README documenting the project
6. Add session info to analysis
7. Make one report parameterized and run with different parameters
8. Share with a colleague - can they run it?

**The ultimate test: Can you run your own analysis 6 months from now?**
