---
title: "Lesson 1: Introduction to R and RMarkdown"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning Objectives

By the end of this lesson, you will be able to:

1. Understand what R and RMarkdown are and why they're useful for biological research
2. Write and execute R code in code chunks
3. Use R as a calculator and understand basic data types
4. Create and manipulate vectors
5. Knit an RMarkdown document to HTML

## What is R?

R is a programming language and environment designed for statistical computing and graphics. It's widely used in biological research for:

- Analyzing experimental data
- Creating publication-quality figures
- Performing statistical tests
- Reproducible research

## What is RMarkdown?

RMarkdown allows you to combine:

- **Narrative text** (like this paragraph)
- **Code** (R commands)
- **Output** (results, figures, tables)

This makes your analysis reproducible and easy to share with collaborators!

## Your First R Code

Let's start simple. R can work as a calculator:

```{r basic-math}
# Addition
5 + 3

# Multiplication
4 * 7

# Division
20 / 4

# Exponentiation (useful for calculating concentrations!)
10^-6  # 1 micromolar in molar
```

**Note:** Lines starting with `#` are comments - R ignores them. Use comments to explain your code!

## Variables: Storing Values

In the lab, you label your tubes. In R, you label your data with variables:

```{r variables}
# Store a single value
sample_concentration <- 2.5

# View the value
sample_concentration

# Use it in calculations
dilution_factor <- 10
final_concentration <- sample_concentration / dilution_factor
final_concentration
```

The `<-` symbol is the assignment operator. Think of it as putting a value into a labeled container.

## Data Types

R has several basic data types:

```{r data-types}
# Numeric (numbers)
cell_count <- 1500000
temperature <- 37.5

# Character (text)
species <- "Escherichia coli"
treatment <- "control"

# Logical (TRUE or FALSE)
is_significant <- TRUE
culture_contaminated <- FALSE

# Check the type
class(cell_count)
class(species)
class(is_significant)
```

## Vectors: Multiple Values

In biology, we rarely work with single measurements. Vectors store multiple values:

```{r vectors}
# Create a vector with c() (combine)
od600_readings <- c(0.1, 0.3, 0.6, 0.9, 1.2)
od600_readings

# Timepoints in hours
timepoints <- c(0, 2, 4, 6, 8)
timepoints

# Character vectors
samples <- c("control", "treatment_A", "treatment_B")
samples
```

## Vector Operations

You can perform calculations on entire vectors at once:

```{r vector-operations}
# Convert OD600 to estimated cell count (approximate)
cells_per_ml <- od600_readings * 8e8
cells_per_ml

# All samples were diluted 1:10
dilution <- 10
actual_concentration <- cells_per_ml * dilution
actual_concentration
```

## Accessing Vector Elements

Use square brackets `[]` to access specific values:

```{r indexing}
# Get the first OD reading
od600_readings[1]

# Get the third and fourth readings
od600_readings[c(3, 4)]

# Get all readings except the first
od600_readings[-1]

# Get readings greater than 0.5
od600_readings[od600_readings > 0.5]
```

## Useful Functions for Vectors

```{r vector-functions}
# How many measurements?
length(od600_readings)

# Summary statistics
mean(od600_readings)
median(od600_readings)
sd(od600_readings)  # standard deviation
min(od600_readings)
max(od600_readings)

# Get a quick summary
summary(od600_readings)
```

## Creating Sequences

Sequences are common in biological work (dilution series, time courses, etc.):

```{r sequences}
# Simple sequence
days <- 1:7
days

# Sequence with specific step
dilutions <- seq(from = 1, to = 0.001, by = -0.1)
dilutions

# Or specify how many values you want
gradient <- seq(from = 0, to = 100, length.out = 11)
gradient
```

## Practice Exercises

Try these on your own:

```{r exercise-1}
# Exercise 1: Create a vector called 'pcr_ct_values' with these Ct values: 18.3, 19.1, 17.8, 20.2, 18.9


# Exercise 2: Calculate the mean and standard deviation of these Ct values


# Exercise 3: You diluted your samples 1:100. Create a variable for this dilution factor


# Exercise 4: Create a vector of sample names: "Sample_1", "Sample_2", "Sample_3", "Sample_4", "Sample_5"


```

## Getting Help

If you need help with a function, use `?`:

```{r help, eval=FALSE}
?mean
?seq
```

## Knitting This Document

To create an HTML report:

1. Click the "Knit" button at the top of this pane
2. Or use Ctrl+Shift+K (Windows/Linux) or Cmd+Shift+K (Mac)

## Key Takeaways

- R is a powerful tool for biological data analysis
- RMarkdown combines code, results, and narrative
- Use `<-` to assign values to variables
- Vectors store multiple values of the same type
- R can perform operations on entire vectors at once
- Comments (`#`) help document your code

## Next Steps

In the next lesson, we'll learn how to import real data and explore it using the tidyverse!

---

**Homework:** Knit this document and experiment with changing some values in the code chunks. What happens if you change the OD600 readings? Try adding your own calculations!
